---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
<!-- badges: end -->

## Calculate Federal and State Income Taxes

`usincometaxes` is an R package that calculates federal and state income taxes in the United States.
It relies on the National Bureau of Economic Research's (NBER) [TAXSIM 32](https://users.nber.org/~taxsim/taxsim32/) 
tax simulator for calculations. The package takes care of the behind-the-scenes work of getting the 
data in the proper format, converting it to the proper file type for uploading to the NBER server, 
uploading the data, downloading the results, and placing the results into a tidy data frame.

*NOTE: This package is not associated with the NBER. It is a private creation that uses their wonderful tax calculator.*

## Installation

You can install `usincometaxes` from [GitHub](https://github.com/shanejorr/usincometaxes) with:

```{r eval = FALSE}
devtools::install_github("shanejorr/usincometaxes")
```

## Quick example

`usincometaxes` helps users estimate household income taxes from data sets containing financial and household data. This allows users to estimate income taxes from surveys with financial information, as the United States Census Public Use Micro Data (PUMS).

`usincometaxes` contains one function: `taxsim_calculate_taxes()`. Below is a simple example of its use.

```{r quickExample}
library(dplyr)
library(knitr)
library(usincometaxes)

family_income <- data.frame(
  id_number = as.integer(c(1, 2)),
  state = c('North Carolina', 'NY'),
  tax_year = c(2015, 2020),
  filing_status = c('single', 'married, jointly'),
  primary_wages = c(10000, 100000),
  primary_age = c(26, 36)
)

family_taxes <- taxsim_calculate_taxes(
  .data = family_income,
  return_all_information = FALSE,
  upload_method = 'ssh'
)
```

```{r}
kable(family_taxes)
```

Users can use the `id_number` column to join the tax data with the original data set. Every `id_number` in
the input data is represented in the output tax data.

```{r joinData}
family_income %>%
  left_join(family_taxes, by = 'id_number') %>%
  kable()
```

## Output

`taxsim_calculate_taxes()` returns a data frame where each row corresponds to a row in `.data` and
each column is a piece of tax information. The output and `.data` can be linked by the `id_number` column.

The amount of output (tax information) received is controlled by the `return_all_information` parameter 
to `taxsim_calculate_taxes()`. Setting `return_all_information` to `FALSE` returns minimal information such as 
federal and state tax liabilities and FICA taxes. When `return_all_information` is `TRUE` 45 different tax 
items are returned.

See the [Description of Output Columns](https://www.shaneorr.io/r/usincometaxes/articles/taxsim-output.html) vignette for output information.

## Input

Taxes are calculated with `taxsim_calculate_taxes()` using the financial and household characteristics 
found in the data frame represented by the `.data` parameter. Each column is a different piece of information
and each row contains a tax payer unit.

All columns must have the column names and data types listed in the 
[Description of Input Columns](https://www.shaneorr.io/r/usincometaxes/articles/taxsim-input.html) vignette.
`.data` can contain columns beyond those listed in the vignette. The additional columns will be ignored.

## Upload and download method

FTP or SSH can be used to upload and retrieve information to and from the TAXSIM server. This is set 
with the `upload_method` parameter to `taxsim_calculate_taxes()` and defaults to FTP. Behind the scenes, 
FTP uses the `RCurl` package and SSH issues an SSH command to the operating system. Large data sets should sue SSH since it is faster.

## Giving credit

The NBER's [TAXSIM 32](https://users.nber.org/~taxsim/taxsim32/) tax simulator does all tax calculations. 
This package simply lets users interact with the tax simulator through R. Therefore, users should 
cite the TAXSIM 32 tax simulator when they use this package in their work:

|           Feenberg, Daniel Richard, and Elizabeth Coutts, An Introduction to the TAXSIM Model, Journal of Policy Analysis and Management vol 12 no 1, Winter 1993, pages 189-194.
