---
title: "Calculating Federal and State Income Taxes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating Federal and State Income Taxes}
  %\VignetteDepends{ggplot2}
  %\VignetteDepends{dplyr}
  %\VignetteDepends{kableExtra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, messages = FALSE}
library(usincometaxes)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggplot2)

# with of scroll box for tables
scroll_width <- "800px"
```

This article presents two use cases for `usincometaxes`. The first shows users how to calculate income taxes from a data frame containing financial information and other characteristics of tax payer units. This income could come from surveys such as the [Consumer Expenditure survey](https://www.bls.gov/cex/) or the [Panel Study of Income Dynamics survey](https://psidonline.isr.umich.edu/). The second use case focuses on running simulations.

For the examples, we will use an internal data set called `taxpayer_finances`. The data is randomly generated, but it is formatted for use with `usincometaxes`.

## Calcualting income taxes from survey data

```{r import_data}
data(taxpayer_finances)

taxpayer_finances %>%
  head() %>%
  kable() %>%
  scroll_box(width = scroll_width)
```

```{r calcualte_survey_taxes}
family_taxes <- taxsim_calculate_taxes(
  .data = taxpayer_finances,
  return_all_information = FALSE,
  upload_method = 'ftp'
)

family_taxes %>%
  head() %>%
  kable()
```

```{r join_tax_data}
income_and_taxes <- taxpayer_finances %>%
  left_join(family_taxes, by = 'id_number')

income_and_taxes %>%
  head() %>%
  kable() %>%
  scroll_box(width = scroll_width)
```



```{r plot_family_taxes}
# custom theme for all plots in the vignette
plt_theme <- function() {
  
    theme_minimal() +
    theme(
      legend.text = element_text(size = 11),
      axis.text = element_text(size = 10),
      axis.title=element_text(size=11,face="bold"),
      strip.text = element_text(size = 11),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(size = 12),
      legend.position = 'bottom'
    )
}

dep_color_palette <- rev(c('#4B0055','#353E7C','#007094','#009B95','#00BE7D','#96D84B'))

# type = dep_color_palette
income_and_taxes %>%
  mutate(
    tax_unit_income = primary_wages + spouse_wages,
    num_dependents_eitc = factor(num_dependents_eitc, levels = as.character(0:5)),
    filing_status = tools::toTitleCase(filing_status)
  ) %>%
  ggplot(aes(tax_unit_income, federal_taxes, color = num_dependents_eitc)) +
  geom_point(alpha = .5) +
  scale_x_continuous(labels = scales::label_dollar(scale = .001, suffix = "K"), limits = c(0, 200000)) +
  scale_y_continuous(labels = scales::label_dollar(scale = .001, suffix = "K"), limits = c(0, 60000)) +
  scale_color_discrete(type = dep_color_palette) +
  facet_grid(rows = vars(filing_status), cols = vars(tax_year)) +
  labs(
    title = "Federal Income Taxes by Filing Status, Year, and Number of Children",
    x = "\nTax Unit Wages",
    y = "Federal Income Taxes"
  ) +
  plt_theme() +
  guides(color = guide_legend(title = "Number of Childern 18 or Younger", title.position = "top", byrow = TRUE))
```

## Income tax simulations

### Income taxes paid by wages

```{r}

# calculate taxes from 0 to 200,000 in wages
wage_linespace <- seq(0, 200000, 100)

n_kids <- 4

base_family_income <- data.frame(
  tax_year = 2020,
  filing_status = 'married, jointly',
  state = 'NC',
  primary_age = 40,
  spouse_age = 40,
  num_dependents = n_kids,
  num_dependents_thirteen = n_kids,
  num_dependents_seventeen = n_kids,
  num_dependents_eitc = n_kids,
  primary_wages = wage_linespace,
  spouse_wages = 0
)

# create an additional data se twith no dependents and add it to the original
family_income <- base_family_income %>%
  bind_rows(
    # make all numeber of dependent columns 0
    base_family_income %>%
      mutate(across(starts_with('num'), ~0))
  ) %>%
  # add unique ID to each row
  mutate(id_number = row_number()) %>%
  select(id_number, everything())

family_income %>%
  head() %>%
  kable() %>%
  scroll_box(width = scroll_width)
```

```{r}
family_income_taxes <- taxsim_calculate_taxes(
  .data = family_income,
  return_all_information = TRUE,
  upload_method = 'ftp'
)

family_income_taxes %>%
  head() %>%
  kable() %>%
  scroll_box(width = scroll_width)
```


```{r}
family_income <- family_income %>%
  left_join(family_income_taxes, by = 'id_number')
```

```{r}
family_income_long <- family_income %>%
  select(primary_wages, num_dependents_eitc, federal_taxes, state_taxes) %>%
  pivot_longer(cols = c('federal_taxes', 'state_taxes'), 
               names_to = 'jurisdiction', values_to = 'taxes_paid') %>%
  mutate(
    jurisdiction = recode(jurisdiction, 'federal_taxes' = 'Federal Taxes', 'state_taxes' = 'NC State Taxes'),
    num_dependents_eitc = factor(num_dependents_eitc, levels = as.character(0:5)),
    post_tax_wages = primary_wages - taxes_paid
  )
# primary_wages, taxes_paid, color = as.character(num_dependents_eitc)
taxes_line_plot <- function(.data, x_var, y_var, color_var) {

  ggplot(.data, aes({{x_var}}, {{y_var}}, color = {{color_var}})) +
    geom_line(size = 1, alpha = .8) +
    geom_hline(yintercept = 0) +
    scale_x_continuous(labels = scales::label_dollar(scale = .001, suffix = "K")) +
    scale_y_continuous(labels = scales::label_dollar(scale = .001, suffix = "K")) +
    scale_color_brewer(type = 'seq', palette = 'Set2')  +
    plt_theme()
  
}

taxes_line_plot(family_income_long, primary_wages, taxes_paid, num_dependents_eitc) +
  facet_wrap(vars(jurisdiction)) +
  labs(
    title = "Relationship Between Wages and Income Taxes Paid",
    subtitle = "Taxpayer is married, filing jointly, in 2020",
    x = "\nTax Unit Wages",
    y = "Federal Income Taxes",
    color = 'Number of Children 18 or Younger:'
  )
```

### Relationship Between Pre and Post-Tax Wages

```{r}

taxes_line_plot(family_income_long, primary_wages, post_tax_wages, num_dependents_eitc) +
  facet_wrap(vars(jurisdiction)) +
  labs(
    title = "Relationship Between Pre and Post-Tax Wages",
    subtitle = "Taxpayer is married, filing jointly, in 2020",
    x = "\nPre-Tax Wages",
    y = "Post-Tax Wages",
    color = 'Number of Children 18 or Younger:'
  )
```

### Federal Earned Income Tax Credit

```{r}
tax_items_mapping <- c(
  eitc = 'Earned Income Tax Credit',
  child_tax_credit = 'Child Tax Credit'
)

family_income %>%
  filter(num_dependents_eitc == 4) %>%
  mutate(child_tax_credit = child_tax_credit_adjusted + child_tax_credit_refundable) %>%
  select(primary_wages, federal_taxes, eitc, child_tax_credit) %>%
  pivot_longer(cols = names(tax_items_mapping), names_to = 'tax_item', values_to = 'amount') %>%
  mutate(tax_item = recode(tax_item, !!!tax_items_mapping)) %>%
  taxes_line_plot(primary_wages, amount, tax_item) +
  labs(
    title = "Relationship Between Wages and Credits",
    subtitle = "Taxpayer is married, filing jointly, in 2020 and has four children under 19",
    x = "\nPre-Tax Wages",
    y = "Credit Amount",
    color = NULL
  )
```
